Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-27 sections/eval.tex
Update on Overleaf.

Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can manipulate an index as a scalar to implement $argmin$. For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
&\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
&\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
&\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}
Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops. In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels. However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.

@@

\subsubsection{Evaluating Finch Logic}
To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) multiple elementwise operations. Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined execution and even takes advantage of vector instructions which are not yet present in Finch.

\begin{figure}
	\includegraphics[width=.3\linewidth]{triangle_count_speedup_over_duckdb.png}
	\includegraphics[width=.35\linewidth]{sddmm_speedup_over_duckdb.png}
	\includegraphics[width=.3\linewidth]{elementwise_speedup_over_duckdb.png}
    \caption{Performance of Finch Logic for common kernels.}
\end{figure}

>>>>

Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can access an index as a scalar to implement $argmin$. For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
&\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
&\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
&\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}
Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops. In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels. However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.

@@

\subsubsection{Evaluating Finch Logic}
To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) multiple elementwise operations. Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined execution and even takes advantage of vector instructions which are not yet present in Finch.
\begin{figure}
\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
  \vspace{0pt} \includegraphics[width=145pt, height=120pt]{figures/elementwise_speedup_over_duckdb.png}
\end{tabular}
\caption{Performance of Finch Logic for common kernels.}
\end{figure}
